A tutorial of Bayesian beta regressions with brms in R

### 2025-02-07 manuscript/manuscript.tex
edit manuscript

Phonetic research often involves numeric continuous outcome variables,
like durations, frequencies, loudness and ratios
\citep{kluender1988, johnson2003, gick2013, ladefoged2014, abramson2017, roettger2018, coretta2023}.
Another commonly employed type of outcome variable are proportions: for
example, proportion of voicing during closure \citep{davidson2016},
vocal folds contact quotient \citep{herbst2017}, gesture amplitude
\citep{carignan2021}, nasalance \citep{carignan2021}. Moreover,
virtually any measure can be MIN-MAX normalised, a procedure which
transforms values so that they are in the range 0--1.

@@

We will use brms to fit a Bayesian beta regression \citep{burkner2017}.
The model has voicing proportion as the outcome variable and the
following terms: an interaction between vowel (/i, e, a, o, u/) and
second consonant C2 (/k, t, p/), centred speech rate (number of
syllables per second); as varying (aka random) terms, we included
by-speaker varying coefficients for the vowel/consonant interaction and
for centred speech rate.\footnote{While the terms ``random effects,
  intercept and slopes'' are commonly employed in applied statistical
  work in linguistics and phonetics, we opt to use the conceptually more
  appropriate terminology proposed by \citet{gelman2005}: varying terms,
  intercepts, slopes, coefficients. Other common terms are multilevel
  and hierarchical terms.} The categorical predictors vowel and C2 are
coded using indexing rather than the traditional R contrasts: in R, this
corresponds to suppressing the model's intercept with the \texttt{0\ +}
syntax; using indexing instead of contrasts makes it easier to specify
priors and uncertainty is distributed equally across all levels of
categorical predictors. For pedagogical simplicity, the model will use
the default priors as explained in Section~\ref{sec-bayes}.

@@

Turning now to the coefficients for vowel and C2, given the indexing
approach of coding these variables the model summary and the output of
\texttt{fixef()} reports the \emph{predictions} in log-odds for each
combination of vowel and C2, rather than differences between levels. The
CrIs of the vowel/C2 coefficients span all negative log-odds values:
these correspond to proportions that are lower than 0.5 (which is 0 in
log-odds). This matches the general trends in the raw data, which we
plotted in Figure~\ref{fig-ita-egg}.

Next, we will plot the predicted proportions of each vowel/C2
combination at mean speech rate (i.e.~centred speech rate = 0) and then
calculate the average pair-wise difference in voicing proportion between
/k, t, p/. Finally, we will assess whether there is greater
between-vowel variability in /t, p/ relative to /k/.

Before being able to plot the predictions, it's important to get
familiar with the so-called posterior draws. {[}Bayesian MCMC XXX{]}.
Posterior draws can be conveniently obtained with the
\texttt{as\_draws\_df()}. For the moment, we will extract only the draws
of the constant regression coefficients (model variables starting with
\texttt{b\_}). To check which coefficients are available in a model, use
\texttt{get\_variables()} from the tidybayes package.
\texttt{as\_draws\_df()} returns a tibble where each column contains the
drawn values of a coefficient. Table~\ref{tbl-voi-prop-draws} shows the
first ten rows and first five columns of the output of
\texttt{as\_draws\_df()}. The probability distribution of the drawn
values of each coefficient is the posterior probability distribution of
that coefficient. Note that, due to the indexing coding of vowel and C2,
all coefficient except \texttt{b\_speech\_rate\_c} are \emph{predicted
log-odds} for each vowel/C2 combination (the drawn values for
\texttt{b\_speech\_rate\_c} are drawn \emph{differences} in log-odds for
each unit increase of speech rate). The drawn values are in log-odds,
but we can convert them to proportions with \texttt{plogis()} (we will
do this when plotting below).

@@

For plotting, we can use ggplot2 statistics layers from the ggdist
package. \texttt{stat\_halfeye()} plots the density of the posterior
probability (in grey), its median (point) and CrIs (lines). Let's use 60
and 80\% CrIs and transform the log-odds values to proportions with
\texttt{plogis()}. See Figure~\ref{fig-voi-prop-bm-1}.

@@

\NormalTok{voi\_prop\_bm\_draws\_long }\SpecialCharTok{|\textgreater{}} 
  \FunctionTok{ggplot}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\FunctionTok{plogis}\NormalTok{(value), vowel)) }\SpecialCharTok{+}
  \FunctionTok{stat\_halfeye}\NormalTok{(}\AttributeTok{.width =} \FunctionTok{c}\NormalTok{(}\FloatTok{0.6}\NormalTok{, }\FloatTok{0.8}\NormalTok{)) }\SpecialCharTok{+}
  \FunctionTok{facet\_grid}\NormalTok{(}\AttributeTok{rows =} \FunctionTok{vars}\NormalTok{(c2))}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{voi\_prop\_bm\_draws\_long\_c2 }\SpecialCharTok{|\textgreater{}} 
  \FunctionTok{ggplot}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\FunctionTok{plogis}\NormalTok{(value\_mean), c2)) }\SpecialCharTok{+}
  \FunctionTok{stat\_halfeye}\NormalTok{(}\AttributeTok{.width =} \FunctionTok{c}\NormalTok{(}\FloatTok{0.6}\NormalTok{, }\FloatTok{0.8}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

Once we have the posterior difference, we can obtain CrIs of the
difference using \texttt{quantile2()} from the posterior package. Beware
that the values of the difference are in log-odds! We can convert these
into odd-ratios with \texttt{exp()}. The output of the code below is
shown in Table~\ref{tbl-voi-bm-quant}. Odd-ratios indicate the ratio of
the difference between A and B, so that 1 means no difference, values
greater than 1 indicate an increase in A relative to B and values lower
than 1 indicate a decrease in A relative to B. odd-ratios are useful
when looking at differences that are in log-odds because while the
relative magnitude of the difference in proportion between two groups is
the same independent of the baseline proportion, the \emph{absolute}
magnitude of the difference depends on the baseline value. For example,
an odd-ratio difference of 1.25 would correspond to a proportion
increase of 13 percentage points if the baseline proportion is 0.62 but
it would correspond to a proportion increase of 17 percentage points if
the baseline proportion is 0.73. Of course, in real research contexts it
is still usefull to think about absolute magnitudes and their relevance
from a conceptual and methodological perspective. In this tutorial we
just focus on odd-ratios for simplicity.

@@

Finally, the package marginaleffects {[}XXX{]} has two other convenience
functions that return CrIs of comparisons across predictor levels
(\texttt{avg\_comparisons()}) and CrIs of posterior predictions across
predictor levels (\texttt{avg\_predictions}). {[}XXX{]}

@@

For the second case study we will use data from \citet{carignan2021}.
The study looked at properties of nasality in German VNC sequences.
Here, we will focus on the effect of C voicing (voiceless /t/ vs voiced
/d/) on the proportion of nasalisation within the vowel in the VNC
sequence. Previous work on coarticulatory nasalisation in English has
suggested that vowels followed by an NC sequence where C is voiceless
(NT) should show earlier coarticulatory nasalisation than vowels
followed by an NC sequence where C is voiced \citep[ND, see review
in][]{carignan2021}. This pattern has been suggested to be driven by the
articulatory and acoustic incompatibility of voicelessness and
nasalisation, by which the velum opening gesture of the nasal consonant
is pushed away (i.e.~earlier) when the consonant following the nasal is
voiceless. Everything else being equal, a greater proportion of vowel
nasalisation (from the perspective of time) should be found in vowels
followed by NT than in vowels followed by ND.

@@

\begin{itemize}
\item
  \texttt{speaker} indicates the speaker ID.
\item
  \texttt{label} is the word label as given in the original data.
\item
  \texttt{vowel} is the target vowel in the VNC sequence.
\item
  \texttt{NC} is the NC sequence.
\item
  \texttt{voicing} indicates the voicing of C.
\item
  \texttt{nas\_prop} is the proportion of coarticulatory nasalisation of
  the vowel.
\end{itemize}

@@

Negative log-odds indicate a proportion that is smaller than 50\%, while
positive log-odds a proportion that is greater than 50\%. Generally, the
expected log-odd predictions are negative, indicating an overall
tendency for the nasalisation to take less than 50\% of the duration of
the vowel. Moreover, the predictions are higher for voiceless NC
sequences than for voiced NC sequences, indicating a greater proportion
of nasalisation in the former. However there is vowel-specific
variation, and there doesn't seem to be much of a difference in
nasalisation proportion in /e/ and /u/. Figure~\ref{fig-nas-prop} shows
the expected predictions with \texttt{conditional\_effects()}, which
should make the interpretation clearer.

@@

To answer question 1, we can calculate the average difference in
nasalisation proportion by first calculating the average nasalisation
across all vowels for voiced and voiceless sequences (see
Table~\ref{tbl-nas-draws} for the output of this step) and then take the
difference of those, similarly to what we have done in the Case Study 1
above.

@@

Now let's calculate the mean nasalisation proportion within each draw by
voicing, and plot the resulting posterior distributions. Note that, as
discussed for Case study 1, when working wit log-odds it is important to
first do all necessary calculations in log-odds, here calculate the mean
log-odds across vowels, and \emph{then} transform the calculated
estimands to proportions/probabilities. The output of the following code
is show in Table~\ref{tbl-nas-long-voi} and the density plot of the
calculated draws is in
Figure~\ref{fig-nas-prop-bm-draws-long-voicing-plot}.

@@

The plot suggests an overall greater nasalisation proportion in
voiceless NC sequences. Let's quantify how greater as we did in Case
Study 1. We will use odd-ratios in this context as well, i.e.~we will
convert log-odds to odd-ratios using the \texttt{exp()} (exponential)
function (and as before we first calculate the difference and then
exponentiate the resulting values, after which we can take summary
measures, like means and quantile-based measures such as CrIs).
tbl-nas-prop-bm-diff-quant show the 90, 80 and 60\% CrIs of the
difference ratio of nasalisation proportion in voiceless vs voiced NC
sequences.

@@

Moving onto question 2: is there individual speaker variation?
{[}TAMMINGA XXX{]} For this, we will use the \texttt{spread\_draws()}
function from tidybayes {[}XXX{]} to extract the draws of the varying
terms (in brms these are the coefficients that start with \texttt{r\_}).
There is quite a few steps of processing to get from the raw draws to
the estimand we need: while we have commented the following code, we
encourage readers to test each line sequentialy and inspect the
intermediate output to fully understand the process. We assume that
readers are familiar enough with models with varying terms (aka random
effects, mixed-effects models). What readers should note is that to
obtain the expected predictions of nasalisation proportion for each
speaker, the constant terms and the varying terms should be added (since
the varying terms indicate the deviation of each speaker from the
overall estimate).

@@

>>>>

Phonetic research often involves numeric continuous outcome variables,
like durations, frequencies, loudness and ratios
\citep{kluender1988, johnson2003, gick2013, ladefoged2014, abramson2017, roettger2018, coretta2023, rosen2005}.
Another commonly employed type of outcome variable are proportions: for
example, proportion of voicing during closure \citep{davidson2016},
vocal folds contact quotient \citep{herbst2017}, gesture amplitude
\citep{carignan2021}, nasalance \citep{carignan2021}. Moreover,
virtually any measure can be MIN-MAX normalised, a procedure which
transforms values so that they are in the range 0--1.

@@

We will use brms to fit a Bayesian beta regression \citep{burkner2017}.
The model has voicing proportion as the outcome variable and the
following terms: an interaction between vowel (/i, e, a, o, u/) and
second consonant C2 (/k, t, p/), centred speech rate (number of
syllables per second); as varying (aka random) terms, we included
by-speaker varying coefficients for the vowel/consonant interaction and
for centred speech rate.\footnote{While the terms ``random effects,
  intercept and slopes'' are commonly employed in applied statistical
  work in linguistics and phonetics, we opt to use the conceptually more
  appropriate terminology proposed by \citet{gelman2005}: varying terms,
  intercepts, slopes, coefficients. Other common terms are multilevel
  and hierarchical terms.} The categorical predictors vowel and C2 are
coded using indexing rather than the traditional R contrasts, as
advocated in \citet{mcelreath2019}: in R, this corresponds to
suppressing the model's intercept with the \texttt{0\ +} syntax; using
indexing instead of contrasts makes it easier to specify priors and
uncertainty is distributed equally across all levels of categorical
predictors. For pedagogical simplicity, the model will use the default
priors as explained in Section~\ref{sec-bayes}.

@@

Turning now to the coefficients for vowel and C2, given the indexing
approach of coding these variables the model summary and the output of
\texttt{fixef()} reports the \emph{predictions} in log-odds for each
combination of vowel and C2, rather than differences between levels. The
CrIs of the vowel/C2 coefficients span all negative log-odds values:
these correspond to proportions that are lower than 0.5 (which is 0 in
log-odds). This matches the general trends in the raw data, which we
plotted in Figure~\ref{fig-ita-egg}. Next, we will plot the predicted
proportions of each vowel/C2 combination at mean speech rate
(i.e.~centred speech rate = 0) and then calculate the average pair-wise
difference in voicing proportion between /k, t, p/.

Before being able to plot the predictions, it's important to get
familiar with the so-called posterior draws. Bayesian regression models
are fitted using Markov Chain Monte Carlo (MCMC) methods, as briefly
explained in Section~\ref{sec-bayes}. The \texttt{brm()} function runs
three MCMC chains with 2000 iterations each by default. These are
sufficient in most cases. The first 1000 iterations of each chain are
used for ``warmup'', where the MCMC algorithm is optimised to find the
posterior probability. The second half of the iterations, 1000 per
chain, are used for sampling from the posterior probability. Each
iteration returns a drawn value for each parameter in the model. The
list of values drawn during the Monte Carlo sampling are called the
``posterior draws''. The posterior draws are used for inference: we can
plot them, summarise them, transform them and combine them in any way to
answer specific questions.

Posterior draws can be conveniently obtained with the
\texttt{as\_draws\_df()} from brms. For the moment, we will extract only
the draws of the constant regression coefficients (model variables
starting with \texttt{b\_}). To check which coefficients are available
in a model, use \texttt{get\_variables()} from the tidybayes package
\citep{kay2019}. \texttt{as\_draws\_df()} returns a tibble where each
column contains the drawn values of a coefficient.
Table~\ref{tbl-voi-prop-draws} shows the first ten rows and first five
columns of the output of \texttt{as\_draws\_df()}. The probability
distribution of the drawn values of each coefficient is the posterior
probability distribution of that coefficient. Note that, due to the
indexing coding of vowel and C2, all coefficient except
\texttt{b\_speech\_rate\_c} are \emph{predicted log-odds} for each
vowel/C2 combination (the drawn values for \texttt{b\_speech\_rate\_c}
are drawn \emph{differences} in log-odds for each unit increase of
speech rate). The drawn values are in log-odds, but we can convert them
to proportions with \texttt{plogis()} (we will do this when plotting
below).

@@

For plotting, we can use ggplot2 statistics layers from the ggdist
package {[}\citet{kay2023}; kay2024{]}. \texttt{stat\_halfeye()} plots
the density of the posterior probability (in grey), its median (point)
and CrIs (lines). Let's use 60 and 80\% CrIs and transform the log-odds
values to proportions with \texttt{plogis()}. See
Figure~\ref{fig-voi-prop-bm-1}.

@@

\NormalTok{voi\_prop\_bm\_draws\_long }\SpecialCharTok{|\textgreater{}} 
  \FunctionTok{ggplot}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\FunctionTok{plogis}\NormalTok{(value), vowel)) }\SpecialCharTok{+}
  \FunctionTok{stat\_halfeye}\NormalTok{(}\AttributeTok{.width =} \FunctionTok{c}\NormalTok{(}\FloatTok{0.6}\NormalTok{, }\FloatTok{0.8}\NormalTok{)) }\SpecialCharTok{+}
  \FunctionTok{facet\_grid}\NormalTok{(}\AttributeTok{rows =} \FunctionTok{vars}\NormalTok{(c2)) }\SpecialCharTok{+}
  \FunctionTok{labs}\NormalTok{(}
    \AttributeTok{caption =} \StringTok{"The points are the medians and the two lines are 60 and 80\% CrIs."}
\NormalTok{  )}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{voi\_prop\_bm\_draws\_long\_c2 }\SpecialCharTok{|\textgreater{}} 
  \FunctionTok{ggplot}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\FunctionTok{plogis}\NormalTok{(value\_mean), c2)) }\SpecialCharTok{+}
  \FunctionTok{stat\_halfeye}\NormalTok{(}\AttributeTok{.width =} \FunctionTok{c}\NormalTok{(}\FloatTok{0.6}\NormalTok{, }\FloatTok{0.8}\NormalTok{)) }\SpecialCharTok{+}
  \FunctionTok{labs}\NormalTok{(}
    \AttributeTok{caption =} \StringTok{"The points are the medians and the two lines are 60 and 80\% CrIs."}
\NormalTok{  )}
\end{Highlighting}
\end{Shaded}

@@

Once we have the posterior difference, we can obtain CrIs of the
difference using \texttt{quantile2()} from the posterior package
\citep{burkner2024}. Beware that the values of the difference are in
log-odds! We can convert these into odd-ratios with \texttt{exp()}. The
output of the code below is shown in Table~\ref{tbl-voi-bm-quant}.
Odd-ratios indicate the ratio of the difference between A and B, so that
1 means no difference, values greater than 1 indicate an increase in A
relative to B and values lower than 1 indicate a decrease in A relative
to B. Odd-ratios are useful when looking at differences that are in
log-odds because while the relative magnitude of the difference in
proportion between two groups is the same independent of the baseline
proportion, the \emph{absolute} magnitude of the difference depends on
the baseline value. For example, an odd-ratio difference of 1.25 would
correspond to a proportion increase of about 3 percentage points if the
baseline proportion is 0.62 but it would correspond to a proportion
increase of 5 percentage points if the baseline proportion is 0.73 and 1
percentage point if the baseline is 0.99. Of course, in real research
contexts it is still useful to think about absolute magnitudes and their
relevance from a conceptual and methodological perspective. In this
tutorial we just focus on odd-ratios for simplicity.

@@

Finally, the package marginaleffects \citep{arelbundock2025} has two
other convenience functions that return CrIs of comparisons across
predictor levels (\texttt{avg\_comparisons()}) and CrIs of posterior
predictions across predictor levels (\texttt{avg\_predictions}). Both
functions take the model and a list of variables as arguments.
Optionally, you can set the confidence level with the
\texttt{conf\_level} argument (here, we set it to 0.8, or 80\%).
\texttt{avg\_comparisons()} allows the user to specify the type of
comparison for each variable: here we set \texttt{"pairwise"} to
calculate pairwise comparisons of each level of \texttt{c2}. The output
of the two functions is shows in Table~\ref{tbl-marginal-1} and
Table~\ref{tbl-marginal-2}.

@@

For the second case study we will use data from \citet{carignan2021}.
The study looked at properties of nasality in German VNC sequences.
Here, we will focus on the effect of C voicing (voiceless /t/ vs voiced
/d/) on the proportion of nasalisation within the vowel in the VNC
sequence. Previous work on coarticulatory nasalisation in English has
suggested that vowels followed by an NC sequence where C is voiceless
(NT) should show earlier coarticulatory nasalisation than vowels
followed by an NC sequence where C is voiced (ND). This pattern has been
suggested to be driven by the articulatory and acoustic incompatibility
of voicelessness and nasalisation, by which the velum opening gesture of
the nasal consonant is pushed away (i.e.~earlier) when the consonant
following the nasal is voiceless
\citep{fowler2005, beddor2009, cho2017, carignan2015}. Everything else
being equal, a greater proportion of vowel nasalisation (from the
perspective of time) should be found in vowels followed by NT than in
vowels followed by ND.

@@

\begin{itemize}
\item
  \texttt{speaker} indicates the speaker ID.
\item
  \texttt{label} is the word label as given in the original data.
\item
  \texttt{vowel} is the target vowel in the VNC sequence.
\item
  \texttt{NC} is the NC sequence (voiceless /nt/ or voiced /nd/).
\item
  \texttt{voicing} indicates the voicing of C.
\item
  \texttt{nas\_prop} is the proportion of coarticulatory nasalisation of
  the vowel.
\end{itemize}

@@

Negative log-odds indicate a proportion that is smaller than 50\%, while
positive log-odds a proportion that is greater than 50\%. Generally, the
expected log-odd predictions in Table~\ref{tbl-nas-prop-fixef} are
negative, indicating an overall tendency for the nasalisation to take
less than 50\% of the duration of the vowel. Moreover, the predictions
are higher for voiceless NC sequences than for voiced NC sequences,
indicating a greater proportion of nasalisation in the former. However
there is vowel-specific variation, and there doesn't seem to be much of
a difference in nasalisation proportion in /e/ and /u/.
Figure~\ref{fig-nas-prop} shows the expected predictions with
\texttt{conditional\_effects()}, which should make the interpretation of
the predictions from Table~\ref{tbl-nas-prop-fixef} clearer.

@@

To answer question 1, we can calculate the average difference in
nasalisation proportion by first calculating the average nasalisation
across all vowels for voiced and voiceless sequences (see
Table~\ref{tbl-nas-draws} for the output of this step) and then take the
difference of those, similarly to what we have done in the case study in
Section~\ref{sec-case-1}.

@@

Now let's calculate the mean nasalisation proportion within each draw by
voicing, and plot the resulting posterior distributions. Note that, as
discussed in Section~\ref{sec-case-1}, when working with log-odds it is
important to first do all necessary calculations in log-odds (here
calculate the mean log-odds across vowels) and \emph{then} transform the
calculated estimands to proportions/probabilities. The output of the
following code is show in Table~\ref{tbl-nas-long-voi} and the density
plot of the calculated draws is in
Figure~\ref{fig-nas-prop-bm-draws-long-voicing-plot}.

@@

The plot suggests an overall greater nasalisation proportion in
voiceless NC sequences. Let's quantify how greater as we did in
Section~\ref{sec-case-1}. We will use odd-ratios in this context as
well, i.e.~we will convert log-odds to odd-ratios using the
\texttt{exp()} (exponential) function (and as before we first calculate
the difference and then exponentiate the resulting values, after which
we can take summary measures, like means and quantile-based measures
such as CrIs). Table~\ref{tbl-nas-prop-bm-diff-quant} show the 90, 80
and 60\% CrIs of the difference ratio of nasalisation proportion in
voiceless vs voiced NC sequences.

@@

Moving onto question 2: is there individual speaker variation? While in
the recent tradition of linguistic research to use varying (aka random)
effects to \emph{control} for differences across participants, it has
been proposed to use the estimates of the varying coefficients to
investigate individual differences \citep{tamminga2016}. We will use the
\texttt{spread\_draws()} function from tidybayes \citep{kay2019} to
extract the draws of the varying terms (in brms these are the
coefficients that start with \texttt{r\_}). There is quite a few steps
of processing to get from the raw draws to the estimand we need: while
we have commented the following code, we encourage readers to test each
line sequentially and inspect the intermediate output to fully
understand the process. We assume that readers are familiar enough with
models with varying terms (aka random effects, mixed-effects models).
What readers should note is that to obtain the expected predictions of
nasalisation proportion for each speaker, the constant terms and the
varying terms should be added (since the varying terms indicate the
deviation of each speaker from the overall estimate).

@@

\section{Conclusion}\label{conclusion}

In this tutorial we have presented two case studies where the outcome
variable is a beta variable, i.e.~a numeric continuous variable bounded
between 0 and 1. We have shown how to fit Bayesian beta regressions to
coarticulatory progressive voicing in voiceless stops in Italian in
Section~\ref{sec-case-1} and coarticulatory regressive nasalisation in
vowels in German in Section~\ref{sec-case-2}. While this tutorial alone
will not be sufficient to be able to independently run full Bayesian
analyses, it will serve as a first introduction for readers to jump
start their learning.
